Apparent preferences for cognitive effort fade when multiple forms of effort

### 2021-05-28 Manuscript.Rmd
exp 1 methods revised

<!-- NOTES: 
- If you're using Rstudio on a mac, you can press 'ctrl + option + o' to compress all chunks and look at the paper + code outline. 
- This script is set so you can run each code snippet independently (after you run the setups and data loading snippets above the institutional affiliations)
-->

@@

  The experimental task was run using PsychoPy 2 [v1.85.1, @Peirce2019] on a Macbook Pro laptop. Participants foraged for monetary rewards in an environment in which each trial required either physical effort or cognitive effort for a set period of time (the “handling time”), or an equivalent unfilled delay (Fig. 1A). Their goal was to maximize their gains within a fixed amount of time. On each trial a monetary offer was displayed for 2 s, and participants had the opportunity to expend time and/or effort during the handling time in order to earn it. Upon completion of a trial, participants saw a 2 s window displaying the reward obtained, which was followed by a travel time to the next offer. Alternatively, the participant could quit at any point during the handling time by pressing the spacebar on the computer, and immediately start traveling.

  Each participant was assigned to one of three types of costs (cognitive effort, physical effort, or delay), and a fourth group of equal size faced an effortless physical task that involved minimal gripping. This latter condition was added to discard the possibility that any differences in acceptances between effort and delay were driven by engagement alone. Each group was unaware that other cost conditions existed. Participants exerted physical effort by gripping a handheld dynamometer (Biopac Systems, United States) using their dominant hand. Gripping requirements were calibrated at 20% of maximum voluntary contraction (MVC, acquired at the beginning of the session). Cognitive effort entailed switching among Stroop, dot motion coherence, and flanker tasks. This type of task switching is demonstrably effortful (Kool et al., 2010). In the Stroop task, one of three color names was displayed on the screen (red, blue or green), with a font color that was either congruent (e.g. the word red presented in red) or incongruent (e.g. the word red presented in blue). Participants had to select the color of the font (i.e., they had to suppress their tendency to read the word). For the motion coherence task, 100 solid white dots moved on the screen. A fraction of these dots moved coherently to the left or right, while the rest moved in random directions (coherence could be either 30% or 40%, uniformly sampled). Participants had to respond with the direction of the coherent movement. In the flanker task, rows of arrowheads pointed either to the left or the right (maximum of 3 rows, 3 to 13 arrowheads per row). Participants responded with the direction of the center arrowhead, which could point in the same or opposite direction from its neighbors. The tasks were configured so that responses always involved a left or right key press (e.g. for Stroop, two colored circles were presented at each side of the screen). During the handling time, cognitive tasks and their configurations were randomly sampled, and were presented for 1 s followed by a 1 s inter-stimulus interval. Participants were asked to respond within each task’s presentation time (i.e., there was a 1 s response deadline). Before the experiment, participants trained in each cognitive task until they correctly performed six consecutive tasks of each kind. While in the handling time, if participants failed to maintain above-threshold gripping or made two mistakes during the cognitive task, they were forced to travel and missed the reward.

  There were three block types, in which handling times of 2, 10, or 14 seconds were paired with travel times of 14, 6, and 2 seconds, respectively. Note that all combinations add up to 16 seconds, meaning that accepted trials took the same amount of time in all conditions, but rejected ones were shorter in environments where the traveling time was shorter. Timing parameters were held constant within each 7-min block. Each block type was presented twice in pseudo-randomized order (total session length = 42 minutes). Reward amounts varied uniformly per trial (4, 8, or 20 cents), with the constraint that every reward was presented twice every six trials. This prevented sequences from being dominated by a single amount during any window of time. Timing information was disclosed at the beginning of each block, and rewards displayed during a 2 s offer window before each trial began. Participants received training prior to the experimental session and were told about all possible timing conditions and reward amounts in order to preclude biases due to experience-dependent learning [@Dundon2020; @Garrett2020].

@@

#### *Operationalization of Cost*

@@

  Foraging theory posits that accepting a delayed reward should depend on the opportunity cost (OC) of time to be incurred in obtaining it, which depends on the richness of the environment [@Charnov1976; @Stephens1986]. In this study, the richness of the environment was manipulated by the length of the handling and travel times, where shorter travel times produced richer environments. Since time combinations were fixed per block, we calculated each block’s optimal accept/reject strategy by computing the expected reward rate (g(*)) from all decision strategies according to the following equation:

$$\begin{aligned} 
g(*) = \dfrac {\sum p_i R_i} { 3\text{Time}_\text{offer} + (\sum p_i H + \text{Time}_\text{reward}) + 3T},p \in \{0, 1\}
\end{aligned}$$

  where *R* and *p* are the reward and acceptance probability of offer *i*, respectively, *H* is the handling time for the block, *T* its associated travel time, $\text{Time}_\text{offer}$ is the duration of the offer window, and $\text{Time}_\text{reward}$ is the duration of the window displaying the reward earned (only presented for completed trials). This gave us the total reward per second attainable in each block as a function of the acceptance threshold (i.e. the lowest amount accepted). Since an ideal forager should always either accept or reject different categories of prey (Krebs et al., 1977), probabilities other than 0 and 1 are not considered. The resulting possible strategies were to accept only 20 cents, accept 8 and 20 cents, or accept everything. Fig. 1B shows possible earnings per second for each choice strategy, as well as the lowest amount participants should accept in order to maximize their rewards (circled dots). For example, a participant in a 10 s handling block should accept 8-cent and 20-cent rewards (and reject 4-cent rewards) to maximize their reward rate. Note that accepting every offer was often detrimental to participant earnings.

@@

  All analyses were performed in R 3.5.2 [@RTeam2018]. First we tested whether decision makers integrated delay and reward information. To address the prediction that participants would be more likely to accept higher reward and shorter handling time trials, we fit a mixed-effects logistic regression to predict trial-wise acceptances [using the lme4 package, @Bates2015], giving a random intercept to each subject to account for biases in participant behavior. We included continuous regressors for handling time and reward, as well as two covariates that probed the influence of recent history on choices. The first was an autoregressive regressor containing the number of consecutive unobtained offers prior to a given trial (unobtained offers could be due to quits or forced travels). The second regressor tracked the sum of the previous *n* offers. We identified the *n* that minimized Akaike's Information Criterion (AIC) among 6 versions of the model (tracing from 2 previous trials up to 7, beyond which the model failed to converge), and reported all coefficients from the winning model.

  Our foraging task was configured such that over- and under-accepting were detrimental to total earnings. To confirm this, we fit a general linear model with constant, linear, and quadratic terms to estimate the correspondence between proportion of trials completed (independent variable) and total earnings (dependent variable). A significant quadratic coefficient thus would signal that the task statistics operated as expected. Next, to determine the optimality of the group’s decisions, we examined whether each cost type produced a bias to over or under-accept offers 4 and 8 cents (assuming that 20 cents were always be accepted, given our design). The reward-maximizing strategy was to always reject these offers for handling times of 14 s, reject 4 cents and accept 8 cents for 10 s, and accept both for 2 s, yielding a combined optimal proportion of acceptances of 50% (see Operationalization of Costs for details). We performed a two-sided one-sample chi-squared test of proportions against the null probability of 0.5 for each type of cost. Therefore, a significant difference indicated that participants either over- or under-accepted rewards. This approach was adapted from the analysis preregistered for Experiment 2, as we found it better suited to the question than the original formulation.

  Next, we compared acceptance rates across cost conditions. We first performed a one-way ANOVA on the proportion of trials accepted using group as a factor. We then compared the proportion completed among all 4 groups using non-parametric permutation contrasts (6 tests). For each test, on each of 5000 permutation iterations, the group assignment was randomly shuffled without replacement, and the difference in mean acceptance rates across iterations created a putative null distribution. The unpermuted group mean difference was then evaluated against this permuted distribution. The same approach was used for total earnings.

  In order to further look at the effect of handling, offer, and cost type, we computed the probability of accepting a trial with a mixed-effects logistic regression. Based on the task structure and our main question, our a priori model of interest included cost condition, handling time, and reward amount as fixed main effects, and a random intercept per subject. Cost condition was modeled with three categorical terms, with the fourth condition as the reference condition. We ran three versions of the model with different reference conditions, in order to test all pairwise differences among the four cost conditions. We then examined whether this a priori model outperformed both simpler and more complex models. We used both AIC and Bayesian Information Criterion (BIC) to determine the model that minimized the negative log-likelihood while penalizing the addition of parameters. The regressors included in each model were: 1) intercept only; 2) condition only; 3) handling time only; 4) offer amount only; 5) handling time and offer amount; 6) condition, handling time, and reward main effects (a priori model from above); 7) adding a handling-by-reward interaction; and 8) all possible two-way interactions. We predicted that model 6 (the a priori setup) would have the lowest AIC and BIC. Nested models with similar AIC were statistically compared using an analysis-of-deviance. The significance test was computed as the probability of the reduction in deviance, based on a chi-square distribution with degrees of freedom equal to the difference in the number of parameters between models.

  Reward-maximizing prey foragers should either accept or reject offers based on their profitability relative to their alternatives [@Krebs1977; @Stephens1986]. However, decision makers often curtail persistence based on continuous reevaluation of delayed prospects or lapses in self control [@Ainslie1975; @McGuire2012; @McGuire2015; @Mischel1970]. Therefore, we examined whether participants were consistent (not quitting a trial after it was originally accepted) and stable (similar acceptance rates over time) in their choices. For consistency, we visually examined survival curves indicating at what point during the handling time participants quit a trial (censored points included trial completions and forced travels), in addition to quitting time distributions. We interrogated stability by computing each participant’s total proportion of acceptances in the first and second half of the experiment for every block type, comparing the mean proportion of acceptances using paired permutations (5000 iterations). We then tested whether the observed differences in costs were still present on each half of the experiment by applying winning model from the mixed-effects logistic comparison to each half separately. In each case, the reference cost category was rotated in order to assess pairwise differences among all costs. We evaluated the consistency of the relative costs by visually comparing the patterns of significant differences between experimental halves.

### **Results**
#### *Between-subjects: Decision Makers Integrate Delay and Reward Information*

@@

temp2 <- apply(temp, 1, function(row) {prop.test(as.numeric(row[3]), sum(as.numeric(row[3:4])), p = 0.5, conf.level = 0.95)})

>>>>

<!-- NOTES: 
- If you're using Rstudio on a mac, you can press 'ctrl + option + o' to compress all chunks and look at the paper + code outline. 
- This script is set so you can run each code snippet independently (after you run the setups and data loading snippets above the institutional affiliations)
- Numbers on each chunk correspond generally to sections in the preregistrations
-->

@@

  The experimental task was run using PsychoPy 2 [v1.85.1, @Peirce2019] on a Macbook Pro laptop. Participants foraged for monetary rewards in an environment in which each trial required physical effort, cognitive effort, or passive waiting for a set period of time (the “handling time”) in order to obtain the reward (Fig. 1A). Their goal was to maximize their gains within a fixed total amount of playing time (six 7-minute blocks). On each trial, a monetary offer was displayed for 2 s, and participants had the opportunity to expend time and/or effort during the handling time in order to earn it. Upon completion of a trial, participants saw a 2 s feedback message displaying the reward obtained, which was followed by a travel time to the next offer. Alternatively, the participant could quit at any point during the handling time by pressing the spacebar, and immediately start traveling. Participants received their earnings in the form of a real monetary bonus that was added to a fixed \$12 participation amount (possible range of task-based earnings = \$0 - \$15.50, rounded up to the nearest quarter).

  Participants were divided into four groups. Each participant in the first three groups was assigned to one of three types of costs (cognitive effort, physical grip-force effort, or delay), and a fourth group experienced a low-effort physical task that required gripping above a negligible force level. This last condition was added to test the possibility that any differences in acceptances between effort and delay were driven by task engagement alone. Each group was unaware that other cost conditions existed. Participants exerted physical effort by gripping a handheld dynamometer (Biopac Systems, United States) using their dominant hand. Gripping requirements in the physical effort condition were calibrated at 20% of maximum voluntary contraction (acquired at the beginning of the session). The low-effort condition was equivalent except that the grip-force threshold was set to 5%. Cognitive effort entailed switching among Stroop, dot motion judgment, and flanker tasks. Task switching has previously been shown to impose subjective costs (Kool et al., 2010). In the Stroop task, one of three color names was displayed on the screen (red, blue or green), with a font color that was either congruent (e.g. the word red presented in red) or incongruent (e.g. the word red presented in blue). Participants had to select the color of the font (i.e., they had to suppress their tendency to read the word). For the motion judgment task, 100 solid white dots moved on the screen. A fraction of these dots moved coherently to the left or right, while the rest moved in random directions (coherence could be either 30% or 40%, uniformly sampled). Participants had to respond with the direction of the coherent movement. In the flanker task, rows of arrowheads pointed either to the left or the right (maximum of 3 rows, 3 to 13 arrowheads per row). Participants responded with the direction of the center arrowhead, which could point in the same or opposite direction of its neighbors. The tasks were configured so that responses always involved a left or right key press (e.g. for Stroop, two colored circles were presented at each side of the screen). During the handling time, cognitive tasks and their configurations were randomly sampled, and were presented for 1 s followed by a 1 s inter-stimulus interval. Participants were asked to respond within each task’s presentation time (i.e., there was a 1 s response deadline). Before the experiment, participants trained in each cognitive task until they correctly performed six consecutive trials of each type. While in the handling time, if participants failed to maintain an above-threshold grip during the physical effort task (following the 1 s grace period) or made two mistakes during the cognitive task, they were forced to travel and missed the reward.

  There were three block types, in which handling times of 2, 10, or 14 seconds were paired with travel times of 14, 6, and 2 seconds, respectively. All combinations added up to 16 seconds, meaning that accepted trials took the same amount of time in all conditions, but rejected trials were shorter in environments with shorter travel times. Timing parameters were held constant within each 7-min block. Each block type was presented twice in pseudorandom order (total session length = 42 minutes). Reward amounts varied per trial (4, 8, or 20 cents with equal probability), with the constraint that every reward was presented twice every six trials. This prevented sequences from being dominated by a single amount during any window of time. Timing information was disclosed at the beginning of each block, and each trial's reward amount was displayed during a 2 s offer window before the trial began. Participants received training prior to the experimental session and were told about all possible timing conditions and reward amounts in order to avoid any biases that could arise through experience-dependent learning [@Dundon2020; @Garrett2020]. Finally, we encouraged participants to carefully evaluate their options by telling them that accepting all offers would not maximize their rewards.

@@

#### *Reward-maximizing Strategy*

@@

  Foraging theory posits that the decision to accept a delayed reward should depend on the opportunity cost of time to be incurred in obtaining it, which in turn depends on the richness of the environment [@Charnov1976; @Stephens1986]. In this study, the richness of the environment was manipulated by the length of the handling and travel times. Shorter travel times produced richer environments by making it possible to save more time by skipping low-reward trials. Since time combinations were fixed per block, we calculated each block’s optimal accept/reject strategy by computing the expected rate of gain (g(*)) from all decision strategies according to the following equation:

$$\begin{aligned} 
g(*) = \dfrac {\sum p_i \text{R}_{i}} { 3\text{T}_\text{offer} + \sum p_i (\text{T}_\text{handling} + \text{T}_\text{reward}) + 3 \text{T}_\text{travel}},p \in \{0, 1\}
\end{aligned}$$

  where *R* and *p* are the reward magnitude and acceptance probability of offer *i*, respectively, $\text{T}_\text{handling}$ is the handling time for the block, $\text{T}_\text{travel}$ is the travel time, $\text{T}_\text{offer}$ is the duration of the offer window (2 s), and $\text{T}_\text{reward}$ is the duration of the window displaying the reward earned (2 s, only presented for completed trials). This gives the average reward per second attainable in each block as a function of the acceptance threshold (i.e. the lowest amount the participant is willing to accept in a given timing environment). Since an ideal forager should always either accept or reject any given category of prey (Krebs et al., 1977), probabilities other than 0 and 1 were not considered. The resulting possible strategies were to accept only 20 cents, accept 8 and 20 cents, or accept everything. Fig. 1B shows possible earnings per second for each choice strategy, as well as the lowest amount participants should accept in order to maximize their rewards (circled dots). For example, a participant in a 10 s handling-time block should accept 8-cent and 20-cent rewards (and reject 4-cent rewards) to maximize their reward rate; a participant in a 14 s handling-time block should accept only 20-cent rewards, and a participant in a 2 s handling-time block should accept everything. Note that accepting every offer was often detrimental to earnings.

@@

  All analyses were performed in R 3.5.2 [@RTeam2018]. First we tested whether decision makers integrated delay and reward information. To address the prediction that participants would be more likely to accept higher reward and shorter handling time trials, we fit a mixed-effects logistic regression to predict trial-wise acceptances [using the lme4 package, @Bates2015], giving a random intercept to each subject to account for biases in participant behavior. We included continuous regressors for handling time and reward, as well as two covariates that probed the influence of recent history on choices. The first was a term containing the number of consecutive unobtained offers prior to a given trial (unobtained offers could be due to quits or forced travels). The second regressor tracked the sum of the reward amounts on the previous *n* offers. We identified the *n* that minimized Akaike's Information Criterion (AIC) among 6 versions of the model (ranging from 2 previous trials up to 7, beyond which the model failed to converge), and report all coefficients from the winning model, which used the sum of the previous 6 offers.

  Our foraging task was configured such that over- and under-accepting were detrimental to total earnings. To confirm this, we fit a general linear model with constant, linear, and quadratic terms to estimate the correspondence between proportion of trials completed (independent variable) and total earnings (dependent variable). A significant quadratic coefficient thus would signal that the task statistics operated generally as expected. Next, to determine the optimality of the group’s decisions, we examined whether each cost type produced a bias to over or under-accept offers of 4 and 8 cents (assuming that 20cent-offers would always be accepted, given our design). The reward-maximizing strategy was to always reject these offers for handling times of 14 s, reject 4 cents and accept 8 cents for 10 s, and accept both for 2 s, yielding a combined optimal overall acceptance rate of 50%. We performed a two-sided one-sample chi-squared test of proportions against the null probability of 0.5 for each type of cost (with trials pooled across participants in each group). Therefore, a significant difference indicated that participants either over- or under-accepted rewards.

  The analyses above differed slightly from the pre-registeered plan. Rather than pooling coefficients from individual logistic regressions, we favored obtaining a single coefficient from a mixed model that accounted for participant-level random effects. The chi-squared test was adopted from the Experiment 2 preregistration, as we found it better suited to the question than the original formulation (i.e. individual tests for each combination of experimental parameters).

  Next, we compared acceptance rates across cost conditions. We first performed a one-way ANOVA on the proportion of trials accepted using group as a factor. We then compared the proportion completed across all 4 groups using non-parametric permutation contrasts (6 tests). For each test, on each of 5000 permutation iterations, participants' group labels were randomly shuffled without replacement, and the difference in mean acceptance rates across iterations created an empirical null distribution. The unpermuted group mean difference was then evaluated against this permuted distribution. The same approach was used to test differences in total earnings.

  In order to further look at the effect of handling time, offer amount, and cost condition, we computed the probability of accepting a trial with a mixed-effects logistic regression. Our a priori model of interest included all three variables as fixed main effects, and a random intercept per subject. Cost condition was modeled with three categorical terms, with the fourth condition as the reference condition. We ran three versions of the model with different reference conditions, in order to test all pairwise differences among the four cost conditions. We then examined whether this a priori model outperformed both simpler and more complex models. We used both AIC and Bayesian Information Criterion (BIC) to determine the model that minimized the negative log-likelihood while penalizing the addition of parameters. The regressors included in the 8 candidate models were: 1) intercept only; 2) cost condition only; 3) handling time only; 4) offer amount only; 5) handling time and offer amount; 6) cost condition, handling time, and offer amount main effects (a priori model from above); 7) adding a handling-time-by-offer-amount interaction; and 8) all possible two-way interactions. We predicted that model 6 would have the lowest AIC and BIC. Nested models with similar AIC were statistically compared using an analysis-of-deviance. The significance test was computed as the probability of the reduction in deviance, based on a chi-square distribution with degrees of freedom equal to the difference in the number of parameters between models.

  Reward-maximizing prey foragers should either accept or reject offers based on their profitability relative to the available alternatives [@Krebs1977; @Stephens1986]. However, decision makers often curtail persistence before reaching a delayed prospect based on a continuous reevaluation of the prospect's value or a lapse in self control [@Ainslie1975; @McGuire2012; @McGuire2015; @Mischel1970]. Therefore, we examined whether participants were consistent (not quitting trials that were originally accepted) and stable (maintaining similar acceptance rates over time) in their choices. To evaluate consistency, we visually examined survival curves indicating at what point during the handling time participants quit a trial; large numbers of late quit times would indicate a lack of consistency. We interrogated stability by computing each participant’s total proportion of acceptances in the first and second half of the experiment for every block type, comparing the mean proportion of acceptances using paired permutations (5000 iterations). We then tested whether the observed differences in costs were still present in each half of the experiment by applying the winning model from the mixed-effects logistic regression analysis to each half separately.

### **Results**
#### *Decision Makers Integrate Delay and Reward Information*

@@

# apply chi-squared
temp2 <- apply(temp, 1, function(row) {prop.test(as.numeric(row[3]), sum(as.numeric(row[3:4])), p = 0.5, conf.level = 0.95)})
